Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-25 paper/analysis.tex
make notation consistent and clean up tex files

Consider a general parameter-dependent matrix
$\mtx{B}(t) \in \mathbb{R}^{n \times n}$ for which each
entry $b_{ij}(t)$ is a 
%\begin{equation*}
%     \mtx{B}(t) = \begin{bmatrix}
%         b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
%         b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
%         \vdots & \vdots & \ddots & \vdots \\
%         b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
%     \end{bmatrix} \in \mathbb{R}^{n \times n},
% \end{equation*}
continuous function on a bounded interval $[a,b]$. In this section, we 
describe and analyze methods that estimate, for several values of $t$, the 
trace $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$
from matrix-vector products of $\mtx{B}(t)$ with \emph{constant} random vectors. When $\mtx{B}$ is constant, our methods reduce to known algorithms. When $\mtx{B}$ is variable, important differences arise from the fact that the random vectors do not depend on $t$. On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations. On the other hand, this significantly complicates the analysis, which will be the main focus of this section.

@@

%Write:
%\begin{itemize}
% \item fast singular value decay, intuitively sufficient to consider a low-rank %approximation
% \item motivated by our application we only consider symmetric positive %semi-definite case here (smoothing kernels are positive) for which Nystroem %approximation is the right choice
%\end{itemize}

%\textcolor{red}{Moved from below:
%For symmetric positive semi-definite matrices whose singular values decay %rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be %significantly better than the Girard-Hutchinson estimate. This is reflected in %the following theorem.}
%
%
%Alternatively, the trace of a symmetric matrix whose singular values decay %quickly can be approximated well by using a Gaussian sketching matrix $\mtx%{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström %approximation \cite{gittens-2013-revisiting-nystrom}

\paragraph{Nyström++ estimator.}
The Nyström estimator~\refequ{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\refequ{equ:nystrom-approximation}:
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)),
    \label{equ:nystrompp-trace-estimator}
\end{equation}
where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices. 
%\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}

@@

At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator. Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach. However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation. Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

%In the remainder of this section, we will derive upper bounds on the error of these estimators.
%
%\begin{table}[ht]
%\centering
%\renewcommand{\arraystretch}{1.2}
%\begin{tabular}{@{}lcc@{}}
%\toprule
%Estimator & Matrix & $m=1600$\\
%\midrule
%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
%\bottomrule
%\end{tabular}
%\end{table}

%as we will see in Section ???.

@@

%\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound 
%    \begin{equation} \label{equ:girard-hutchinson-bound}
%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
%    \end{equation}
%    holds with probability at least $1 - \gamma^{-p}$.
%    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds 
%    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
%\end{theorem}

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
\begin{equation}
    \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \Big( \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \|\mtx{B}(t)\|_{F,1} + \frac{2 p}{n_{\mtx{\Psi}}} \|\mtx{B}(t)\|_{2,1} \Big).
    \label{equ:girard-hutchinson-bound}
\end{equation}
In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \|\mtx{B}(t)\|_{F,1}$ holds 
    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
\end{theorem}
\begin{proof}
    Let us first consider $t$ fixed. From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable. Therefore, \reflem{lem:sub-gamma-moments} yields
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 4  \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2.
        \label{equ:moment-bound-hutchinson}
    \end{equation}
%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality. We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
%    \begin{equation}
%        \widetilde{\mtx{B}}(t)
%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
%            \mtx{B}(t) & & \\
%            & \ddots & \\
%            & & \mtx{B}(t)
%        \end{pmatrix}
%        \quad \text{and} \quad
%        \widetilde{\vct{\psi}} = \begin{pmatrix}
%            \vct{\psi}_1 \\
%            \vdots \\
%            \vct{\psi}_{n_{\mtx{\Psi}}}
%        \end{pmatrix}.
%    \end{equation}
%    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$. Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
%    \begin{equation}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
%    \end{equation}
%    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
%    \begin{equation}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
%        \label{equ:moment-bound-hutchinson}
%    \end{equation}

@@

    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$, setting $\gamma = e$, and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows by taking,  e.g., $n_{\mtx{\Psi}} = \lceil (8 e)^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
\end{proof}
%\begin{proof}
%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality. We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
%    \begin{equation}
%        \widetilde{\mtx{B}}(t)
%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
%            \mtx{B}(t) & & \\
%            & \ddots & \\
%            & & \mtx{B}(t)
%        \end{pmatrix}
%        \quad \text{and} \quad
%        \widetilde{\vct{\psi}} = \begin{pmatrix}
%            \vct{\psi}_1 \\
%            \vdots \\
%            \vct{\psi}_{n_{\mtx{\Psi}}}
%        \end{pmatrix}.
%    \end{equation}
%    Note that $\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}} = \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Therefore, by \reflem{lem:hanson-wright-moments} we have
%    \begin{align}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))]
%        &= \mathbb{E}^{p}[\Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}] \notag \\
%        &\leq c \max\left\{ \sqrt{p} \lVert \widetilde{\mtx{B}}(t) \rVert _F, p \lVert \widetilde{\mtx{B}}(t) \rVert _2 \right\} \notag \\
%        &= c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F, \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right\} \notag \\
%        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}} \right\}  \lVert \mtx{B}(t) \rVert _F.
%        \label{equ:moment-bound-hanson-wright}
%    \end{align}
%
%    To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable. Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:moment-bound-hanson-wright} that
%    \begin{align}
%        \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
%        &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
%        &\leq c \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t .
%    \end{align}
%    Consequently, by Markov's inequality,  the inequality 
%    \begin{equation} 
%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
%    \end{equation}
%    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-p}$. 
%
%    Setting $\gamma = e$ and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil c^2 e^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$ and restricting $\varepsilon \leq c e$.
%    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
%    %we get that for all $\delta \in (0, 1)$
%\end{proof}

@@

%The idea behind the proof is to bound  to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
%\begin{proof} 
%\textcolor{red}{??? I feel that the block diagonal embedding trick should come first, like in the proof of Meyer et al. Hopefully, this gets rid of the additional $\log \delta$ factor ???}
%The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality. We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$: 
%    \begin{equation}
%        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
%    \end{equation}
%%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
%for a Gaussian random vector $\vct{\psi}$.
%%
%%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
%The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
%    \begin{align*}
%        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
%        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
%        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
%        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
%    \end{align*}
%%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
%%     \begin{equation}
%%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
%%     \end{equation}
%    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. \textcolor{red}{??? this norm should be defined more prominently somewhere else Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ ??? } \textcolor{green}{Moved it to start of section.} Consequently, the high order moments of $r(\mtx{B}, \vct{\psi})$ are bounded by
%    \begin{equation}
%        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
%        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
%        \label{equ:hutchinson-trace-onequery-fixed}
%    \end{equation}
%    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.
%
%    To address the parameter-dependent case, we first note that the continuity assumption implies that $r(\mtx{B}(t), \vct{\psi})$ is measurable. Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from~\refequ{equ:hutchinson-trace-onequery-fixed} that
%    \begin{equation}
%        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
%        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
%        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
%    \end{equation}
%    Consequently, by Markov's inequality,  the inequality 
%    \begin{equation} 
%        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
%        \label{equ:hutchinson-trace-onequery-uniform}
%    \end{equation}
%    holds for any $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$. 
%
%    To extend the bound~\refequ{equ:hutchinson-trace-onequery-uniform} from $1$ to $n_{\mtx{\Psi}}$ queries, we use a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}). Letting
%    \begin{equation}
%        \widetilde{\mtx{B}}(t)
%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
%            \mtx{B}(t) & & \\
%            & \ddots & \\
%            & & \mtx{B}(t)
%        \end{pmatrix}
%        \quad \text{and} \quad
%        \widetilde{\vct{\psi}} = \begin{pmatrix}
%            \vct{\psi}_1 \\
%            \vdots \\
%            \vct{\psi}_{n_{\mtx{\Psi}}}
%        \end{pmatrix},
%    \end{equation}
%    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Thus, the claimed probabilistic bound~\refequ{equ:girard-hutchinson-bound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.
%%     Then 
%%     where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$, and
%%     \begin{equation}
%%         r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}}) = \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j.
%%     \end{equation}
%%     Hence, by applying \refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ we conclude that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
%%     \begin{equation}
%%         \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
%%         \leq 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
%%     \end{equation}
%
%    %Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) %\rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest %integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac%{1}{2}})$
%    %\begin{equation}
%    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log%(\delta^{-1}) \rceil}}
%    %    = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}%{\lceil \log(\delta^{-1}) \rceil}}
%    %    \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
%    %\end{equation}
%    %from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^%{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
%
%    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
%    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
%    %we get that for all $\delta \in (0, 1)$
%    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
%    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
%    %Consequently, with probability $1 - \delta$ we have
%    %\begin{equation}
%    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
%    %\end{equation}
%\end{proof}
%
%Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the number of queries $n_{\mtx{\Psi}}$ needed by~\refthm{thm:hutchinson} is roughly a factor $\log(\delta^{-1})$ larger.
%
\subsubsection{Nyström estimator}
\label{subsec:nystrom}

@@

Our analysis mimics the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-1/2}$ for the parameter-dependent Nyström approximation.

>>>>

Consider a general parameter-dependent matrix
$\mtx{B}(t) \in \mathbb{R}^{n \times n}$ for which each
entry $b_{ij}(t)$ is a 
continuous function on a bounded interval $[a,b]$. In this section, we 
describe and analyze methods that estimate, for several values of $t$, the 
trace $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$
from matrix-vector products of $\mtx{B}(t)$ with \emph{constant} random vectors. When $\mtx{B}$ is constant, our methods reduce to known algorithms. When $\mtx{B}$ is variable, important differences arise from the fact that the random vectors do not depend on $t$. On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations. On the other hand, this significantly complicates the analysis, which will be the main focus of this section.

@@

\paragraph{Nyström++ estimator.}
The Nyström estimator~\refequ{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\refequ{equ:nystrom-approximation}:
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)),
    \label{equ:nystrompp-trace-estimator}
\end{equation}
where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices.

@@

At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator. Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach. However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation. Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.

@@

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
\begin{equation}
    \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \Big( \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \|\mtx{B}(t)\|_{F,1} + \frac{2 p}{n_{\mtx{\Psi}}} \|\mtx{B}(t)\|_{2,1} \Big).
    \label{equ:girard-hutchinson-bound}
\end{equation}
In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \|\mtx{B}(t)\|_{F,1}$ holds 
    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
\end{theorem}
\begin{proof}
    Let us first consider $t$ fixed. From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable. Therefore, \reflem{lem:sub-gamma-moments} yields
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 4  \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2.
        \label{equ:moment-bound-hutchinson}
    \end{equation}

@@

    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$, setting $\gamma = e$, and choosing $p = \log(\delta^{-1}) \geq 1$, the second part of the theorem follows by taking,  e.g., $n_{\mtx{\Psi}} = \lceil (8 e)^2 \log(\delta^{-1}) \varepsilon^{-2} \rceil$.
\end{proof}

@@

\subsubsection{Nyström estimator}
\label{subsec:nystrom}

@@

Our analysis mimics the strategy from~\cite{meyer-2021-hutch-optimal}, by first establishing a relative Frobenius norm error proportional to $n_{\mtx{\Omega}}^{-\sfrac{1}{2}}$ for the parameter-dependent Nyström approximation.
